---
title: "Bayesian-Based Fantasy Football Prediction Model Proposal"
author: 'STAT 578, Fall 2017, **Team 5**: Aaron Ray, Kiomars Nassiri, Michael Chan'
date: "October 25, 2017"
output:
  pdf_document:
  html_document:
    highlight: haddock
    theme: spacelab
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80)
```

# Project Description

National Football League (NFL), being one of the major professional sports leagues in North America, has a wide audience. One segment of the audience follows NFL in the context of Fantasy Football. Participants in a **[Fantasy Football](https://en.wikipedia.org/wiki/Fantasy_football_(American))** game act as the managers of a virtual footbal team and try to maximize their points by picking up the best line-up. Points are given based on actual performance of players in real-world competition. 
For the purpose of this project we have chosen to work with the data gathered from the **[FanDuel](https://www.fanduel.com/)** internet company. We will use Bayesian hierarchical modeling techniques to predict the FanDuel points of different NFL players. The goal is to predict the FanDuel point of each player in the next game given explanatory variables and provide a posterior credible interval.
The analysis is conducted based on the idea presented in the article, **[Bayesian Hierarchical Modeling Applied to Fantasy Football Projections for Increased Insight and Confidence](http://srome.github.io/Bayesian-Hierarchical-Modeling-Applied-to-Fantasy-Football-Projections-for-Increased-Insight-and-Confidence/)**, by Scott Rome.

# Team Members

- **Aaron Ray** (aaronwr2@illinois.edu)$^*$
- **Kiomars Nassiri** (nassiri2@illinois.edu)
- **Michael Chan** (mhchan3@illinois.edu)

$^*$Contact Person

# Dataset Description

Team has set up a process to gather the historical data from the [RotoGuru](http://rotoguru1.com/cgi-bin/fstats.cgi?pos=0&sort=4&game=f&colA=0&daypt=0&xavg=0&inact=0&maxprc=99999&outcsv=0) website. The following is the code used to get the data from RotoGuru:

```{r, message=FALSE, warning=FALSE,eval=FALSE}
# Scrape rotoguru1 site for weekly FanDuel stats and bind each week's data to the 
# pre-defined dataframe, 'd'.

for(year in 2014:2017){
  for(week in 1:16){
    page = read_html(
      gsub(" ","",
           paste("http://rotoguru1.com/cgi-bin/fyday.pl?week=",week,"&year=",
                 year ,"&game=fd&scsv=1")
      ))
    dtext = page %>% html_nodes("pre") %>% html_text(trim = TRUE)
    dtable = read.table(text=dtext, sep = ";", header=TRUE, col.names = cnames,
                        quote=NULL)
    d = rbind(d,dtable)
  }
}
```

Data cleaning is performed using Excel and R routine. Some data cleaning tasks are needed to calculate Player rank.

## Response Variables

- `FanDuelPts`: Points gained by the player

## Predictor Variables

- `6GmAvgOppPAP`: The six game average Opposing Points Allowed to Position (OppPAP) by the current player's opposing defense. For example, if the Buffalo Bills defense allowed a total of 30 points per game to wide receivers for six games straight, then this number would equal to the average of 30 for any wide receiver facing the Bills defense.
- `Position`: The position the player plays
- `Opponent`: The team that the player plays against
- `Rank`: The rank of a player based on recent performance

# Analysis Ideas

At the lowest level, we model the performance (`FanDuelPts`) as normally-distributed around a true value:

$y|\alpha, \beta_{defense}, \beta_{home}, \beta_{away}, \sigma_r^2 \sim N(\alpha + X_{defense} . \beta_{defense} + X_{home} . \beta_{home} + X_{away} . \beta_{away}, \sigma_y^2 I)$

where

$\alpha$ = `6GmAvgOppPAP`

$\beta_{defense,t,p}$ = defense coefficient against team t for position p

$\beta_{home,p,r}$ = home coefficient for position p and a rank r player

$\beta_{away,p,r}$ = Away coefficient for position p and a rank r player

$y$ = `FanDuelPts`

At higher level, we model the defense effect, $\beta_{defense}$, as how good a player is when playing against a particular team.  We pool the effect based on the position of the player.  That is, the defense coefficient is normally distributed from the same position-specific distribution.

$\beta_{defense,t,p} \sim N(\delta_p, \sigma_{\delta}^2)$

For the home and away game effect, $\beta_{home}$ and $\beta_{away}$, we model the effect for player of the same rank has the same distribution.

$\beta_{home,r} \sim N(\eta_r, \sigma_{\eta}^2)$

$\beta_{away,r} \sim N(\rho_r, \sigma_{\rho}^2)$

We will approximate non informative prior using:

$\sigma_y \sim Inv-gamma(0.0001, 0.0001)$

$\sigma_{\delta} \sim N(0, 100^2)$

$\sigma_{\eta} \sim N(0, 100^2)$

$\sigma_{\rho} \sim N(0, 100^2)$
