---
title: 'STAT 578: Bayesian Hierarchical Modeling Final Project Proposal'
author: "Aaron Ray, Kiomars Nassiri Kahnamooee, Michael Chan, Mohammad Pezeshki"
date: 'Due: Oct 25, 11:59 PM PDT'
output:
  pdf_document:
    toc: yes
  html_document:
    theme: readable
    toc: yes
---

# Final Project Proposal

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80)
```

# Introduction

## Project Members

- Aaron Ray (aaronwr2@illinois.edu)
- Kiomars Nassiri Kahnamooee (nassiri2@illinois.edu)
- Michael Chan (mhchan3@illinois.edu)
- Mohammad Pezeshki (mp13@illinois.edu)

Contact Person: Aaron Ray


## Proposed title for the project

**A Bayesian Statistical Approach to Predicting Fantasy Football**

## Project Description

National Football League (NFL), being one of the major professional sports leagues in North America, has a wide audience. One segment of the audience follows NHL in the context of Fantasy Football.  In this project, we will use Bayesian Hierarchical technique to predict the Fan Duel Points of a NHL player.  The goal is to predict the fan duel point of each player in the next game given explanatory variables and provides a posterior credible interval.  

The analysis is based on the idea presented in the article, Bayesian Hierarchical Modeling Applied to Fantasy Football Projections for Increased Insight and Confidence, by Scott Rome.  [http://srome.github.io/Bayesian-Hierarchical-Modeling-Applied-to-Fantasy-Football-Projections-for-Increased-Insight-and-Confidence/]  


## Description of Dataset

The source of the data for this project: http://rotoguru1.com/cgi-bin/fstats.cgi?pos=0&sort=4&game=f&colA=0&daypt=0&xavg=0&inact=0&maxprc=99999&outcsv=0

Data cleaning is performed using Excel and R routine.  Some data cleaning tasks are needed to calculate Player rank.


### Potential Response Variables

- `FanDuelPts`           Position in final standings

### Predictor Variables

- `6GmAvgOppPAP`         The six game average Opposing Points Allowed to Position (OppPAP) by the current player's opposing defense. For example, if the Buffalo Bills defense allowed a total of 30 points per game to wide receivers for six games straight, then this number would equal to the average of 30 for any wide receiver facing the Bills defense.

- `Position`         The position the player plays
- `Opponent`         The team that the player plays against
- `Rank`             The rank of a player based on recent performance

## Analysis Ideas

At the loweest level, we model the performance(`FanDuelPts`) as normally distributed around a true value.  The model is:


$y|\alpha, \beta_{defense}, \beta_{home}, \beta_{away}, \sigma_r^2 \sim N(\alpha + X_{defense} . \beta_{defense} + X_{home} . \beta_{home} + X_{away} . \beta_{away}, \sigma_y^2 I)$

where

$\alpha$ = `6GmAvgOppPAP`

$\beta_{defense,t,p}$ = defense coefficient against team t for position p

$\beta_{home,p,r}$ = home coefficient for position p and a rank r player

$\beta_{away,p,r}$ = Away coefficient for position p and a rank r player

$y$ = `FanDuelPts`

$x_{t,p}$ = interaction indicator term for team t, position p

$x_{p,r}$ = interaction indicator term for rank r, position p


At higher level, we model the defense effect, $\beta_{defense}$, as how good a player is when playing against a particular team.  We pool the effect based on the position of the player.  That is, the defense coefficient is normally distributed from the same position specific distribution.

$\beta_{defense,t,p} \sim N(\delta_p, \sigma_{\delta}^2)$

For the home and away game effect, $\beta_{home}$ and $\beta_{away}$, we model the effect for player of the same rank has the same distribution.

$\beta_{home,r} \sim N(\eta_r, \sigma_{\eta}^2)$

$\beta_{away,r} \sim N(\rho_r, \sigma_{\rho}^2)$

We will approximate non informative prior using:

$\sigma_y \sim Inv-gamma(0.0001, 0.0001)$

$\sigma_{\delta} \sim N(0, 100^2)$

$\sigma_{\eta} \sim N(0, 100^2)$

$\sigma_{\rho} \sim N(0, 100^2)$

Here is the JAGS model:

```{r eval=FALSE}
model {
  for (i in 1:length(y)) {
    y[i] ~ dnorm(inprod(X.defense[i, ], beta.defense) 
                 + inprod(X.home[i, ], beta.home) 
                 + inprod(X.away[i, ], beta.away), sigmasqinv)
  }
  
  # The entry of the beta.defense corresponds to Opponent:Position
  # In our model, we pool the beta.defense based on position. 
  # i.e. All defense effects of the same position are drawn from the same distribution
  for (p in 1:Num.Position) {
    for (t in 1:Num.Opponent) {
      beta.defense[(p-1) * Num.Opponent + t] ~ dnorm(delta[p], 1/100^2)
    }
    delta[p] ~ dnorm(0, 1/100^2)
  }
  
  # The entry of the beta.home and beta.away corresponds to Rank:Position
  # In our model, we pool the beta.home/away based on rank
  for (r in 1:Num.Rank) {
    for (t in 1:Num.Position) {
      beta.home[(r-1) * Num.Position + t] ~ dnorm(eta[r], 1/100^2)
      beta.away[(r-1) * Num.Position + t] ~ dnorm(rho[r], 1/100^2)
    }
    eta[r] ~ dnorm(0, 1/100^2)
    rho[r] ~ dnorm(0, 1/100^2)
  }  
  
  sigmasqinv ~ dgamma(0.0001, 0.0001)
  sigmasq <- 1/sigmasqinv
}

```


## Sample Data

```{r}
fdp <- read.csv("fdp.csv", sep = '\t', header = TRUE)

head(fdp)
```

*Derived data*
```{r}
#Rank player based on current FanDuelPts 

#(This is cheating as it is looking at future, we'll fix it after more data
# clean up work - This is done only for getting some data to play with in this
# preminlinary research)

year_week = unique(fdp[,c('Year','Week')])
rank_column = "FanDuelPts"
fdp['Rank'] = NA
for (i in 1:nrow(year_week)) {
  fdp_year_week = fdp[fdp$Year == year_week[i, 'Year'] & fdp$Week == year_week[i, 'Week'], ]
  fdp_year_week_quantile = quantile(fdp_year_week[rank_column], c(0.25, 0.5, 0.75), na.rm = TRUE)
  
  fdp[fdp$Year == year_week[i, 'Year'] & fdp$Week == year_week[i, 'Week'] 
      & fdp[rank_column] < fdp_year_week_quantile[1], 'Rank'] = 'Rank4'
  fdp[fdp$Year == year_week[i, 'Year'] & fdp$Week == year_week[i, 'Week'] 
      & fdp[rank_column] >= fdp_year_week_quantile[1] 
      & fdp[rank_column] < fdp_year_week_quantile[2], 'Rank'] = 'Rank3'
  fdp[fdp$Year == year_week[i, 'Year'] & fdp$Week == year_week[i, 'Week'] 
      & fdp[rank_column] >= fdp_year_week_quantile[2] 
      & fdp[rank_column] < fdp_year_week_quantile[3], 'Rank'] = 'Rank2'
  fdp[fdp$Year == year_week[i, 'Year'] & fdp$Week == year_week[i, 'Week'] 
      & fdp[rank_column] >= fdp_year_week_quantile[3], 'Rank'] = 'Rank1'
}
```

```{r}
fdp['Locality'] = 'Away'
fdp[fdp$Home.Game == 1, 'Locality'] = 'Home'

```



** The following R code is for reference and preliminary research **
```{r}
fdp_train=fdp[fdp$Year == 2016, ]

X.defense = model.matrix(~ 0 + Opponent:Position , data=fdp_train)
X.home = model.matrix(~ 0 + Rank:Position , data=fdp_train)
X.away = model.matrix(~ 0 + Rank:Position , data=fdp_train)

X = cbind(X.defense, X.home, X.away)

Num.Opponent = length(unique(fdp[, "Opponent"]))
Num.Position = length(unique(fdp[, "Position"]))
Num.Rank = length(unique(fdp[, "Rank"]))
```

```{r warning=FALSE, message=FALSE}
library(rjags)
set.seed(20171008)

```

```{r}
# Initialization List for the 4 chains
jags.inits=list(
  list( sigmasqinv=    0.01, 
        .RNG.name = "base::Mersenne-Twister", .RNG.seed = 20171008 ),
  list( sigmasqinv=    0.01, 
        .RNG.name = "base::Mersenne-Twister", .RNG.seed = 20171008 + 1 ),
  list( sigmasqinv=0.000001, 
        .RNG.name = "base::Mersenne-Twister", .RNG.seed = 20171008 + 2 ),
  list( sigmasqinv=0.000001, 
        .RNG.name = "base::Mersenne-Twister", .RNG.seed = 20171008 + 3 )
)

data.jags <- list(
  y= fdp_train$FanDuelPts,
  X.defense = X.defense,
  X.home = X.home,
  X.away = X.away,
  Num.Position=Num.Position,
  Num.Opponent=Num.Opponent,
  Num.Rank=Num.Rank
)
```


```{r warning=FALSE}
m <- jags.model("fdp.bug", data.jags, inits = jags.inits, n.chains=4, n.adapt = 1000)
```

```{r warning=FALSE,eval=FALSE}
update(m, 2500) # burn-in

x <- coda.samples(m1, c("delta","beta.defense", "sigmasq"), n.iter=5000)
```


