---
title: "A Bayesian Approach to Predicting NFL Quarterback Scores in Fanduel Tournaments"
author: 'STAT 578, Fall 2017, **Team 5**: Aaron Ray, Kiomars Nassiri, Michael Chan'
date: "December 8, 2017"
output:
  pdf_document:
  html_document:
    highlight: haddock
    theme: spacelab
---
```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80)
```

# Purpose

The National Football League, NFL, is a professional American football league consisting of 32 teams. It is the most professional American football league in the world and is one of the most followed sports leagues in North America. Being a football fan, however, does not end with following matches on TV or stadiums or buying your favorite team's jersey. Over the years, football fans have come up with interesting NFL-themed side-activities, the most popular of which is called "Fantasy Football".

Participants in a **[Fantasy Football](https://en.wikipedia.org/wiki/Fantasy_football_(American))** game act as the managers of a virtual footbal team and try to maximize their points by picking up the best line-up. Points are given based on actual performance of players in real-world competition.Each team is allowed a pre-determined number of players on its roster, as well as a specified number at each position that can or must be used in each game (the "starters").

For the purpose of this project, we have chosen to work with the data gathered from the **[FanDuel](https://www.fanduel.com/)** internet company. FanDuel is a web-based fantasy sports game and with 6 million registered users and is the second largest daily fantasy sports company (as measured by entry fees and user base) in the daily fantasy sports industry.

We will leverage a Hierarchical Bayesian approach with the Markov Chain Monte Carlo method to predict the Fantasy points likely to be scored by an NFL quarterback in any given game.  The goal is to predict the points scored by each player given certain prior conditions and predictor variables that will assist our model in providing credible posterior prediction intervals. The following are the research questions that will be answered in this project:

- **Interpretation**: What features can help us effectively predict Fanduel points players receive in a future match?
- **Prediction**: How reliable are the predictions for the future performance of players?

# Data

Historical data on the performance of the players is extracted from the **[RotoGuru](http://rotoguru1.com/cgi-bin/fstats.cgi?pos=0&sort=4&game=f&colA=0&daypt=0&xavg=0&inact=0&maxprc=99999&outcsv=0)** website. Data scraped from RotoGuru includes information about the Fanduel points received by each player, player's position, player's opponent team, Home/Away match indication, etc. for each week.

The following is the code used to get the data from RotoGuru:

```{r, message=FALSE, warning=FALSE,eval=FALSE}

# Scrape rotoguru1 site for weekly FanDuel stats and bind each week's data to the 
# pre-defined dataframe, 'd'.

for(year in 2014:2017){
  for(week in 1:16){
    page = read_html(
      gsub(" ","",
           paste("http://rotoguru1.com/cgi-bin/fyday.pl?week=",week,"&year=",
                 year ,"&game=fd&scsv=1")
      ))
    dtext = page %>% html_nodes("pre") %>% html_text(trim = TRUE)
    dtable = read.table(text=dtext, sep = ";", header=TRUE, col.names = cnames,
                        quote=NULL)
    d = rbind(d,dtable)
  }
}
```

Data cleaning is performed using R routines. Some data cleaning tasks are needed to calculate Player rank.

## Response Variables

- `FanDuelPts`: Points position at the end of a single game

## Predictor Variables

- `AvgPts5Wks`: The 5 game average points of the player
- `AvgOppPAP7Wks` : The 7 game average Opposing Points Allowed to Position (OppPAP) by the current player's opposing defense. For example, if the Buffalo Bills defense allowed a total of 30 points per game to wide receivers for six games straight, then this number would equal to the average of 30 for any wide receiver facing the Bills defense.
- `Position`: The position the player plays
- `HomeGame`: Whether it is home game.
- `Rank`: The rank of a player based on recent performance

## Sample Data

```{r}
fdp <- read.csv("fdpfinal.csv", sep = ',', header = TRUE)

head(fdp)
```


```{r}
fdp['Rank'] = fdp$OffRnk5Wks
```


## Simple LM graph

$y|\alpha \sim N(\alpha,\sigma_y^2 I)$

```{r}
mod.classic = lm(FanDuelPts ~ AvgPts5Wks, data = fdp)

plot(FanDuelPts ~ AvgPts5Wks, data = fdp)
```

As can be seen from the graph above, the FanDuel data points are all over the place. The model proposed in the next section will present our approach to use the predictors.

\newpage
# Model

At the lowest level, we model the performance (`FanDuelPts`) as normally-distributed around a true value:

$y|\alpha, \beta_{defense}, \beta_{home}, \beta_{away}, \sigma_r^2 \sim N(\alpha + X_{defense} . \beta_{defense} + X_{home} . \beta_{home} + X_{away} . \beta_{away}, \sigma_y^2 I)$

where

$\alpha$ = The average fan duel point of the previous 5 weeks of the player, `AvgPts5Wks`

$\beta_{defense,p}$ = defense coefficient against team t for position p

$\beta_{home,p,r}$ = home coefficient for position p and a rank r player

$\beta_{away,p,r}$ = Away coefficient for position p and a rank r player

$y$ = `FanDuelPts`

$x_{p}$ = interaction indicator term for opposing team score allowed by position p

$x_{home,p,r}$ = interaction indicator term for rank r, position p, and whether it is home game

At the higher level, we model the defense effect, $\beta_{defense}$, as how well a particular team's defense has performed against the player's position. We pool the effect based on the position of the player. That is, the defense coefficient is normally distributed from the same position specific distribution.

$\beta_{defense,p} \sim N(\delta_p, \sigma_{\delta}^2)$

where $\sigma_{\delta}$ is constant = 1000

For the home and away game effect, $\beta_{home}$ and $\beta_{away}$, we model the effect for player of the same rank has the same distribution.  We model the home and away game effect to be the same for players of the same position.

$\beta_{home,p,r} \sim N(\eta_r, \sigma_{\eta}^2)$

$\beta_{away,p,r} \sim N(\rho_r, \sigma_{\rho}^2)$

where $\sigma_{\eta}$, $\sigma_{\rho}$ are constant = 1000

We will approximate non informative prior using:

$\sigma_y \sim Inv-gamma(0.0001, 0.0001)$

$\delta \sim N(0, 10000^2)$

$\eta \sim N(0, 10000^2)$

$\rho \sim N(0, 10000^2)$

Here is the JAGS model:

```{r eval=FALSE}
#sink("fdp.bug")
#cat("
model {
  for (i in 1:length(y)) {
    y[i] ~ dnorm(alpha[i] + inprod(X.defense[i, ], beta.defense) 
                 + inprod(X.home[i, ], beta.home) 
                 + inprod(X.away[i, ], beta.away), sigmasqinv)
  }
  
  # The entry of the beta.defense corresponds to Opponent:Position
  # In our model, we pool the beta.defense based on position. 
  # i.e. All defense effects of the same position are drawn from the same distribution
  for (p in 1:Num.Position) {
    for (f in 1:Num.fixed.pred) {
      beta.defense[(f-1) * Num.Position + p] ~ dnorm(delta[p], 1/1000^2)
      delta[(f-1) * Num.Position + p] ~ dnorm(0, 1/100000^2)
    }
  }
  
  # The entry of the beta.home and beta.away corresponds to Rank:Position
  # In our model, we pool the beta.home/away based on rank
  for (r in 1:Num.Rank) {
    for (t in 1:Num.Position) {
      beta.home[(t-1) * Num.Rank + r] ~ dnorm(eta[r], 1/1000^2)
      beta.away[(t-1) * Num.Rank + r] ~ dnorm(rho[r], 1/1000^2)
    }
    eta[r] ~ dnorm(0, 1/100000^2)
    rho[r] ~ dnorm(0, 1/100000^2)
  }

  sigmasqinv ~ dgamma(0.0001, 0.0001)
  sigmasq <- 1/sigmasqinv
}
#    ",fill = TRUE)
#sink()

```

```{r echo=FALSE, message=FALSE, warning=FALSE, r echo=FALSE}
library(knitr)
```

\newpage
# Computation

We use overdispersed starting points and 4 chains to initalized.  We have seen slow mix in and hence we use thinning of 5 to get a smaller number of data points.  Convergence diagnostics are performed graphically as well as using Gelman Statistics.  We make sure sample size are > 400.  The Morte Carlo error of the $\beta$ are less then 0.06, but the ones of the hyper parameters are much higher at around 5.

## Training Data Setup

Due to significant roster change usually happens in the offseason, we believe it is the best to not use data across seasons.  We use 2016 data as it is the most recent year with a full reason. We set aside the last week for verifying prediction effectiveness for the model.

```{r}
#fdp_train=fdp[fdp$Year == 2015, ]
fdp_train=fdp[fdp$Year == 2016  & ((fdp$Position == "QB"& fdp$FanDuelSalary > 6500 & !is.na(fdp$FanDuelSalary))|fdp$Position == "PK"), ]
#fdp_train=fdp[fdp$Year == 2015 & (fdp$Position == "QB" | fdp$Position == "RB") & fdp$FanDuelSalary > 6500 & !is.na(fdp$FanDuelSalary), ]

fdp_test =fdp_train[fdp_train$YearWeek >= 201615, ]
fdp_test = droplevels(fdp_test)
fdp_train=fdp_train[fdp_train$YearWeek < 201615, ]
fdp_train = droplevels(fdp_train)
```

```{r}
Use.Rank = TRUE
Num.Opponent = length(unique(fdp_train[, "Opponent"]))
Num.Position = length(unique(fdp_train[, "Position"]))
#Num.fixed.pred=2 #AvgOppPAP7Wks + FanDuelSalary
Num.fixed.pred=1 #AvgOppPAP7Wks
if (Use.Rank) {
  Num.Rank = length(unique(fdp_train[, "Rank"]))
  Num.HomeAwayInit = Num.Rank
  Model.File.Ext = ""
} else {
  Num.HomeAwayInit = 1
  Model.File.Ext = ".norank"
}


if (Num.Position == 1) {
  #X.defense = model.matrix(~ 0 + AvgOppPAP7Wks + FanDuelSalary, data=fdp_train)
  X.defense = model.matrix(~ 0 + AvgOppPAP7Wks, data=fdp_train)
  if (Use.Rank) {
    X.home = model.matrix(~ 0 + Rank , data=fdp_train)
    X.away = model.matrix(~ 0 + Rank , data=fdp_train)
  } else {
    X.home = rep(1, nrow(fdp_train))
    X.away = rep(1, nrow(fdp_train))
  }
} else {
  #X.defense = model.matrix(~ 0 + AvgOppPAP7Wks:Position + FanDuelSalary:Position, data=fdp_train)
  X.defense = model.matrix(~ 0 + AvgOppPAP7Wks:Position, data=fdp_train)
  if (Use.Rank) {
    X.home = model.matrix(~ 0 + Rank:Position , data=fdp_train)
    X.away = model.matrix(~ 0 + Rank:Position , data=fdp_train)
  } else {
    X.home = model.matrix(~ 0 + Position , data=fdp_train)
    X.away = model.matrix(~ 0 + Position , data=fdp_train)
  }
}


X.home = X.home * fdp_train$HomeGame
X.away = X.away * (1- fdp_train$HomeGame)
X = cbind(X.defense, X.home, X.away)

```

The following code set up the data used in the model computation.  `X.defense` captures the interaction terms between the defensive power (higher `AvgOppPAP7Wks` means the player is facing a weaker team, since they allows players score more point on them) and position.

**Sample X.defense**
```{r}
head(X.defense)
```

`X.home` and `X.away` are the interaction terms between Position and Rank.  Together `X.home` and `X.away` will always sum to one.  The intercept is implicitly included.

**Sample X.home**
```{r}
head(X.home)
```



```{r warning=FALSE, message=FALSE, echo=FALSE}
library(rjags)
set.seed(20171008)

```

##Initialization

```{r}
# Initialization List for the 4 chains
jags.inits=list(
  list( sigmasqinv=    0.01,  delta = rep(-100000, Num.Position * Num.fixed.pred),
        eta = c(100000, -100000, 100000, -100000)[1:Num.HomeAwayInit],
        rho = c(-100000, 100000, -100000, 100000)[1:Num.HomeAwayInit],
        .RNG.name = "base::Mersenne-Twister", .RNG.seed = 20171008 ),
  list( sigmasqinv=    0.01,  delta = rep(100000, Num.Position * Num.fixed.pred),
        eta = c(100000, -100000, -100000, 100000)[1:Num.HomeAwayInit],
        rho = c(-100000, 100000, 100000, -100000)[1:Num.HomeAwayInit],
        .RNG.name = "base::Mersenne-Twister", .RNG.seed = 20171008 + 1 ),
  list( sigmasqinv=0.000001,  delta = rep(-100000, Num.Position * Num.fixed.pred),
        eta = c(-100000, 100000, -100000, 100000)[1:Num.HomeAwayInit],
        rho = c(100000, -100000, 100000, -100000)[1:Num.HomeAwayInit],
        .RNG.name = "base::Mersenne-Twister", .RNG.seed = 20171008 + 2 ),
  list( sigmasqinv=0.000001,  delta = rep(100000, Num.Position * Num.fixed.pred),
        eta = c(-100000, 100000, 100000, -100000)[1:Num.HomeAwayInit],
        rho = c(100000, -100000, -100000, 100000)[1:Num.HomeAwayInit],
        .RNG.name = "base::Mersenne-Twister", .RNG.seed = 20171008 + 3 )
)

data.jags <- list(
  y= fdp_train$FanDuelPts,
  alpha = fdp_train$AvgPts5Wks,
  X.defense = X.defense,
  X.home = X.home,
  X.away = X.away,
  Num.fixed.pred=Num.fixed.pred,
  Num.Position=Num.Position,
  #Num.Opponent=Num.Opponent,
  Num.Rank=Num.Rank
)
```



```{r warning=FALSE}
burnAndSample = function(m, N.burnin, N.iter, show.plot, mon.col, n.thin=1) {
  update(m, N.burnin) # burn-in
  
  x <- coda.samples(m, mon.col, n.iter=N.iter, n.thin)
  
  if(show.plot) {
    plot(x, smooth=FALSE)
  }
  
  gelman.R = gelman.diag(x, autoburnin=FALSE, multivariate = FALSE)

  result <- list(
    coda.sam = x, 
    gelman.R.max=max(gelman.R$psrf[, 1]),
    gelman.R = gelman.R
  )
  
  return(result)
}

```

```{r warning=FALSE}
runModel=TRUE
runSample=TRUE

mon.col <- c("delta", "eta", "rho", "beta.defense", "beta.home", "beta.away", "sigmasq")

NSim = 30000
NChain = 4
NThin = 5
NTotalSim = NSim * NChain / 5
if (runModel) {
  m <- jags.model("fdp.bug", data.jags, inits = jags.inits, n.chains=NChain, n.adapt = 1000)
  save(file=paste("fdp.jags.model.init", Model.File.Ext, ".Rdata", sep=""), list="m")
} else {
  load(paste("fdp.jags.model.init", Model.File.Ext, ".Rdata", sep=""))
  m$recompile()
}

load.module("dic")

N.Retry.Loop = 1
if (runSample) {
  N.burnin=2500/2
  for (loopIdx in 1:N.Retry.Loop) {
    (start_time <- Sys.time())
    (N.burnin = N.burnin * 2)
    result = burnAndSample(m, N.burnin, NSim, show.plot=FALSE, mon.col = mon.col, n.thin=NChain)
    (end_time <- Sys.time())
    (result$gelman.R.max)    
  }
  run.params = paste(".", N.burnin, ".", NChain, ".", NSim, ".", NThin, sep="")
  save(file=paste("fdp.jags.samples", run.params, Model.File.Ext, ".Rdata", sep=""), list="result")
  save(file=paste("fdp.jags.model", run.params, Model.File.Ext, ".Rdata", sep=""), list="m")
} else {
  N.burnin=2500/2 * (2**N.Retry.Loop)
  run.params = paste(".", N.burnin, ".", NChain, ".", NSim, ".", NThin, sep="")
  load(paste("fdp.jags.samples", run.params, Model.File.Ext, ".Rdata", sep=""))
  load(paste("fdp.jags.model", run.params, Model.File.Ext, ".Rdata", sep=""))
  m$recompile()
}
```

##Convergence diagnostics

###Trace Plots

The trace plots of all parameters show good distribution convergence.

```{r fig.width=10,fig.height=11}
plot(result$coda.sam, smooth=FALSE)
```

###Gelman Statistics

```{r}
result$gelman.R
```

Converged as `gelman.R.max` = `r result$gelman.R.max` < 1.1 and the plot also looks good.

###MCMC Summary
```{r}
(m.summary = summary(result$coda.sam))
```


###Effective Sample Size
```{r}
(eff.size = effectiveSize(result$coda.sam[, ]))
```

The effective sample sizes of all parameters are greater than 400.

\newpage
#Model Assessment

##General posterior model assumption check

*Probabiliy of players should perform better at home than away*

```{r}
post.samp = as.matrix(result$coda.sam)

beta.home = post.samp[, paste("beta.home[",1:(Num.Rank*Num.Position),"]", sep="")]
beta.away = post.samp[, paste("beta.away[",1:(Num.Rank*Num.Position),"]", sep="")]

prob.home.away = rep(0, Num.Rank * Num.Position)
for (r in 1:Num.Rank) {
  for (p in 1:Num.Position) {
    idx = (p-1) * Num.Rank + r
    prob.home.away[idx] = mean(beta.home[, idx] > beta.away[, idx])
  }
}
prob.home.away
prob.home.away.df = data.frame(colnames(X.home))
prob.home.away.df$prob.home.bt.away = prob.home.away

colnames(prob.home.away.df) = c("Rank:Position", "Prob.home.bt.away")

kable(prob.home.away.df)

```

The above table shows that players perform better at home than away as expected.

*Beta defense*

If a player is facing a team which gives up more points to players on average, we expect the player will score more points.
```{r}
Num.fixed.size=Num.Position*Num.fixed.pred
beta.defense = post.samp[, paste("beta.defense[",1:Num.fixed.size,"]", sep="")]

beta.defense.int.df = data.frame(colnames(X.defense))
beta.defense.int = matrix(rep(0, Num.fixed.size * 4), nrow=Num.fixed.size, ncol = 4)

int.alpha=0.05
for (p in 1:Num.Position) {
  for (f in 1:Num.fixed.pred) {
    idx = (f-1) * Num.Position + p
      beta.defense.int[idx, 1:3] =  quantile(beta.defense[, idx], c(int.alpha/2, 0.5, 1-int.alpha/2))
      beta.defense.int[idx, 4] = mean(beta.defense[, idx])
  }
}

beta.defense.int.df$pct025 = beta.defense.int[, 1]
beta.defense.int.df$pct975 = beta.defense.int[, 3]
beta.defense.int.df$median = beta.defense.int[, 2]
beta.defense.int.df$mean = beta.defense.int[, 4]
colnames(beta.defense.int.df) = c("beta.defense.position", "pct025", "pct975", "median", "mean")
kable(beta.defense.int.df)


```

We observe that the median beta.defense for PK is positive as expected.  But for QB, it is negative, that implies QB actually scores less against bad defensive team.  

*DIC*
```{r}
(dic.samp = dic.samples(m, NTotalSim))

```

The effective number of parameters("penalty") is 19, and the Plummer's DIC ("Penalized deviance") is 4719.  Not that we have `r ncol(post.samp)` parameters in our model, `r ncol(post.samp) -19` of them were shrunk away.

## Posterior Predictive Check

### Error correlation Check
A posterior predictive p-value usng the following test quantity

$$
\begin{aligned}
T(y, X, \theta ) & = |\hat{cor}(\epsilon, \text{time})|  \\
\end{aligned}
$$

where $\hat{cor}(\epsilon, \text{time})$ is sample correlation between the error vector $\epsilon$ and the year week in the data. The larger this quantity is for the model, the less well it fits the data as that would mean the error is correlated with time.

*The simulated error vectors $\epsilon$ (as rows of a matrix):*
```{r}
error.sim <- matrix(NA, NTotalSim, nrow(fdp_train))
y_hat.sim <- matrix(NA, NTotalSim, nrow(fdp_train))
for(s in 1:NTotalSim) {
  y_hat.sim[s, ] =  fdp_train$AvgPts5Wks + (X.defense %*% beta.defense[s, ]) + (X.home %*% beta.home[s, ]) + (X.away %*% beta.away[s, ])
  error.sim[s, ] <- (fdp_train$FanDuelPts - y_hat.sim[s, ]) 
}

```

*The simulated replicate error vectors $\epsilon^{rep}$ (as rows of a matrix), which are the error vectors computed using replicate response vectors $y^{rep}$:*

```{r}

post.sigma.2.sim <- post.samp[,"sigmasq"]
post.sigma.sim <- sqrt(post.sigma.2.sim)

yreps <- matrix(NA, NTotalSim, nrow(fdp_train))
for(s in 1:NTotalSim) {
  yreps[s, ] <- rnorm(nrow(fdp_train), y_hat.sim[s, ], post.sigma.sim[s])
}

error.rep <- matrix(NA, NTotalSim, nrow(fdp_train))

for(s in 1:NTotalSim) {
  error.rep[s, ] <- (yreps[s, ] - y_hat.sim[s, ]) 
}

```

*The simulated values of $T(y, X, \theta )$*
```{r}
T.sim = abs(cor(t(error.sim), fdp_train$YearWeek))
head(T.sim)
```

*The simulated values of $T(y^{rep}, X, \theta )$*
```{r}
T.rep.sim = abs(cor(t(error.rep), fdp_train$YearWeek))
head(T.rep.sim)
```

*The simulated values of $T(y^{rep}, X, \theta )$ versus those of  $T(y, X, \theta )$, with a reference line indicating where the two values would be equal.*

```{r}
plot(T.sim, T.rep.sim, pch=".", cex=2,
  xlim=c(min(T.sim, T.rep.sim), max(T.sim, T.rep.sim)),
  ylim=c(min(T.sim, T.rep.sim), max(T.sim, T.rep.sim)),
  xlab="T(y,x,theta)", ylab="T(y.rep,x,theta)")
abline(a=0,b=1)
```
The posterior predictive p-value:

```{r}
(p.value = mean(T.rep.sim >= T.sim))
```

The p.value is `r p.value`, > 0.05, which does not indicate any evidence of problem.

### Chi-square Discrepancy Check

```{r}


Tchi <- numeric(NTotalSim)
Tchirep <- numeric(NTotalSim)
for(s in 1:NTotalSim){
  Tchi[s] <- sum((fdp_train$FanDuelPts - y_hat.sim[s,])^2 / post.sigma.sim[s])
  Tchirep[s] <- sum((yreps[s,] - y_hat.sim[s,])^2 / post.sigma.sim[s])
}
(p.value.Tchi = mean(Tchirep >= Tchi))
```

```{r}
plot(Tchi, Tchirep, pch=".", cex=2,
  xlim=c(min(Tchi, Tchirep), max(Tchi, Tchirep)),
  ylim=c(min(Tchi, Tchirep), max(Tchi, Tchirep)),
  xlab="Tchi", ylab="Tchirep")
abline(a=0,b=1)
```

The posterior predictive p-value using the chi-square discrepancy is `p.value.Tchi`=`r p.value.Tchi`.  The p-value is > 0.05.  Hence, it does not indicate any evidence of problems.

### Individual Data Point Discrepancy Check

**Using $Pr(y^{rep} \geq y | y)$ as posterior predictive p-value**
```{r}
yreps.minus.y <- matrix(NA, NTotalSim, nrow(fdp_train))
for(s in 1:NTotalSim) {
  yreps.minus.y[s, ] <- yreps[s, ] - fdp_train$FanDuelPts
}

(p.value.y.rep.all = mean(yreps.minus.y > 0))
```

The posterior predictive p-value using individual data point is `p.value.y.rep.all` = `r p.value.y.rep.all`, which is > 0.05.  This shows no evidence of problem.


## Prediction

WE use the last week of data to check the prediction effectiveness of the model

```{r}
if (Num.Position == 1) {
  #X.defense = model.matrix(~ 0 + AvgOppPAP7Wks + FanDuelSalary, data=fdp_train)
  X.defense.test = model.matrix(~ 0 + AvgOppPAP7Wks, data=fdp_test)
  if (Use.Rank) {
    X.home.test = model.matrix(~ 0 + Rank , data=fdp_test)
    X.away.test = model.matrix(~ 0 + Rank , data=fdp_test)
  } else {
    X.home.test = rep(1, nrow(fdp_test))
    X.away.test = rep(1, nrow(fdp_test))
  }
} else {
  #X.defense = model.matrix(~ 0 + AvgOppPAP7Wks:Position + FanDuelSalary:Position, data=fdp_train)
  X.defense.test = model.matrix(~ 0 + AvgOppPAP7Wks:Position, data=fdp_test)
  if (Use.Rank) {
    X.home.test = model.matrix(~ 0 + Rank:Position , data=fdp_test)
    X.away.test = model.matrix(~ 0 + Rank:Position , data=fdp_test)
  } else {
    X.home.test = model.matrix(~ 0 + Position , data=fdp_test)
    X.away.test = model.matrix(~ 0 + Position , data=fdp_test)
  }
}


X.home.test = X.home.test * fdp_test$HomeGame
X.away.test = X.away.test * (1- fdp_test$HomeGame)
X.test = cbind(X.defense.test, X.home.test, X.away.test)
```

```{r}
y_hat.test <- matrix(NA, NTotalSim, nrow(fdp_test))
for(s in 1:NTotalSim) {
  y_hat.test[s, ] =  fdp_test$AvgPts5Wks + (X.defense.test %*% beta.defense[s, ]) + (X.home.test %*% beta.home[s, ]) + (X.away.test %*% beta.away[s, ])
}

y.pred <- matrix(NA, NTotalSim, nrow(fdp_test))
for(s in 1:NTotalSim) {
  y.pred[s, ] <- rnorm(nrow(fdp_test), y_hat.test[s, ], post.sigma.sim[s])
}

```

*Prediction of indvidual data point*

Let's pick one example data point, `fdp_test[1,]`

```{r}
(fdp_test[1,])
```

The real FanDuelPts is `r fdp_test$FanDuelPts[1]`

The prediction value has the following 95% interval

```{r}
quantile(y.pred[, 1], c(0.025, 0.975))
```

which does contain the actual data value of `r fdp_test$FanDuelPts[1]`


Let's look at $Pr(y_{pred} >= y)$

```{r}
y.pred.minus.y <- matrix(NA, NTotalSim, nrow(fdp_test))
for(s in 1:NTotalSim) {
  y.pred.minus.y[s, ] <- y.pred[s, ] - fdp_test$FanDuelPts
}

(p.value.y.pred.all = mean(y.pred.minus.y > 0))
```

The probability of $y_{pred} >= y$ is `r p.value.y.pred.all`, close to 0.5, which implies a relatively good predictive value.  

*A look at a cross section of how one simulation of a prediction of the whole test set*

```{r}
for (s in 1:1) {
  plot(fdp_test$FanDuelPts, y.pred[s, ], pch=".", cex=2,
  xlim=c(min(y.pred[s, ], fdp_test$FanDuelPts), max(y.pred[s, ], fdp_test$FanDuelPts)),
  ylim=c(min(y.pred[s, ], fdp_test$FanDuelPts), max(y.pred[s, ], fdp_test$FanDuelPts)),
  xlab="y", ylab="y pred")
  abline(a=0,b=1)
}

```

\newpage
##Alternative Model - no rank

```{r eval=FALSE}
#sink("fdp.norank.bug")
#cat("
model {
  for (i in 1:length(y)) {
    y[i] ~ dnorm(alpha[i] + inprod(X.defense[i, ], beta.defense) 
                 + inprod(X.home[i, ], beta.home) 
                 + inprod(X.away[i, ], beta.away), sigmasqinv)
  }
  
  # The entry of the beta.defense corresponds to Opponent:Position
  # In our model, we pool the beta.defense based on position. 
  # i.e. All defense effects of the same position are drawn from the same distribution
  for (p in 1:Num.Position) {
    beta.defense[p] ~ dnorm(delta[p], 1/1000^2)
    delta[p] ~ dnorm(0, 1/100000^2)
  }
  
  # The entry of the beta.home and beta.away corresponds to Position
  # In our model, we pool the beta.home/away based on Position
  # NO RANK
  for (t in 1:Num.Position) {
    beta.home[t] ~ dnorm(eta, 1/1000^2)
    beta.away[t] ~ dnorm(rho, 1/1000^2)
  }
  eta ~ dnorm(0, 1/100000^2)
  rho ~ dnorm(0, 1/100000^2)

  sigmasqinv ~ dgamma(0.0001, 0.0001)
  sigmasq <- 1/sigmasqinv
}
#    ",fill = TRUE)
#sink()

```

```{r}
Use.Rank = FALSE
Num.Opponent = length(unique(fdp_train[, "Opponent"]))
Num.Position = length(unique(fdp_train[, "Position"]))
#Num.fixed.pred=2 #AvgOppPAP7Wks + FanDuelSalary
Num.fixed.pred=1 #AvgOppPAP7Wks
if (Use.Rank) {
  Num.Rank = length(unique(fdp_train[, "Rank"]))
  Num.HomeAwayInit = Num.Rank
  Model.File.Ext = ""
} else {
  Num.HomeAwayInit = 1
  Model.File.Ext = ".norank"
}


if (Num.Position == 1) {
  #X.defense = model.matrix(~ 0 + AvgOppPAP7Wks + FanDuelSalary, data=fdp_train)
  X.defense = model.matrix(~ 0 + AvgOppPAP7Wks, data=fdp_train)
  if (Use.Rank) {
    X.home = model.matrix(~ 0 + Rank , data=fdp_train)
    X.away = model.matrix(~ 0 + Rank , data=fdp_train)
  } else {
    X.home = rep(1, nrow(fdp_train))
    X.away = rep(1, nrow(fdp_train))
  }
} else {
  #X.defense = model.matrix(~ 0 + AvgOppPAP7Wks:Position + FanDuelSalary:Position, data=fdp_train)
  X.defense = model.matrix(~ 0 + AvgOppPAP7Wks:Position, data=fdp_train)
  if (Use.Rank) {
    X.home = model.matrix(~ 0 + Rank:Position , data=fdp_train)
    X.away = model.matrix(~ 0 + Rank:Position , data=fdp_train)
  } else {
    X.home = model.matrix(~ 0 + Position , data=fdp_train)
    X.away = model.matrix(~ 0 + Position , data=fdp_train)
  }
}


X.home = X.home * fdp_train$HomeGame
X.away = X.away * (1- fdp_train$HomeGame)
X = cbind(X.defense, X.home, X.away)

```

```{r}
# Initialization List for the 4 chains
jags.inits=list(
  list( sigmasqinv=    0.01,  delta = rep(-100000, Num.Position * Num.fixed.pred),
        eta = c(100000, -100000, 100000, -100000)[1:Num.HomeAwayInit],
        rho = c(-100000, 100000, -100000, 100000)[1:Num.HomeAwayInit],
        .RNG.name = "base::Mersenne-Twister", .RNG.seed = 20171008 ),
  list( sigmasqinv=    0.01,  delta = rep(100000, Num.Position * Num.fixed.pred),
        eta = c(100000, -100000, -100000, 100000)[1:Num.HomeAwayInit],
        rho = c(-100000, 100000, 100000, -100000)[1:Num.HomeAwayInit],
        .RNG.name = "base::Mersenne-Twister", .RNG.seed = 20171008 + 1 ),
  list( sigmasqinv=0.000001,  delta = rep(-100000, Num.Position * Num.fixed.pred),
        eta = c(-100000, 100000, -100000, 100000)[1:Num.HomeAwayInit],
        rho = c(100000, -100000, 100000, -100000)[1:Num.HomeAwayInit],
        .RNG.name = "base::Mersenne-Twister", .RNG.seed = 20171008 + 2 ),
  list( sigmasqinv=0.000001,  delta = rep(100000, Num.Position * Num.fixed.pred),
        eta = c(-100000, 100000, 100000, -100000)[1:Num.HomeAwayInit],
        rho = c(100000, -100000, -100000, 100000)[1:Num.HomeAwayInit],
        .RNG.name = "base::Mersenne-Twister", .RNG.seed = 20171008 + 3 )
)

data.jags <- list(
  y= fdp_train$FanDuelPts,
  alpha = fdp_train$AvgPts5Wks,
  X.defense = X.defense,
  X.home = X.home,
  X.away = X.away,
  Num.fixed.pred=Num.fixed.pred,
  Num.Position=Num.Position
  #Num.Opponent=Num.Opponent,
  #Num.Rank=Num.Rank
)
```

```{r warning=FALSE}
runModel=TRUE
runSample=TRUE

mon.col <- c("delta", "eta", "rho", "beta.defense", "beta.home", "beta.away", "sigmasq")

NSim = 30000
NChain = 4
NThin = 5
NTotalSim = NSim * NChain / 5
if (runModel) {
  if(Use.Rank) {
    bug.file = "fdp.bug"
  } else {
    bug.file = "fdp.norank.bug"
  }
  m <- jags.model(bug.file, data.jags, inits = jags.inits, n.chains=NChain, n.adapt = 1000)
  save(file=paste("fdp.jags.model.init", Model.File.Ext, ".Rdata", sep=""), list="m")
} else {
  load(paste("fdp.jags.model.init", Model.File.Ext, ".Rdata", sep=""))
  m$recompile()
}

load.module("dic")

N.Retry.Loop = 1
if (runSample) {
  N.burnin=2500/2
  for (loopIdx in 1:N.Retry.Loop) {
    (start_time <- Sys.time())
    (N.burnin = N.burnin * 2)
    result = burnAndSample(m, N.burnin, NSim, show.plot=FALSE, mon.col = mon.col, n.thin=NChain)
    (end_time <- Sys.time())
    (result$gelman.R.max)    
  }
  run.params = paste(".", N.burnin, ".", NChain, ".", NSim, ".", NThin, sep="")
  save(file=paste("fdp.jags.samples", run.params, Model.File.Ext, ".Rdata", sep=""), list="result")
  save(file=paste("fdp.jags.model", run.params, Model.File.Ext, ".Rdata", sep=""), list="m")
} else {
  N.burnin=2500/2 * (2**N.Retry.Loop)
  run.params = paste(".", N.burnin, ".", NChain, ".", NSim, ".", NThin, sep="")
  load(paste("fdp.jags.samples", run.params, Model.File.Ext, ".Rdata", sep=""))
  load(paste("fdp.jags.model", run.params, Model.File.Ext, ".Rdata", sep=""))
  m$recompile()
  gelman.diag(result$coda.sam, autoburnin=FALSE, multivariate = FALSE)
}
```

Converged as `gelman.R.max` = `r result$gelman.R.max` < 1.1 and the plot also looks good.

```{r}
(m.summary = summary(result$coda.sam))
```


*Effective Sample Size*
```{r}
(eff.size = effectiveSize(result$coda.sam[, ]))
```
The effective sample sizes of all parameters are greater than 400.

*DIC*
```{r}
(dic.samp = dic.samples(m, NTotalSim))

```

The effective number of parameters("penalty") is 7.02, and the Plummer's DIC ("Penalized deviance") is 4781.  This model has a higher DIC compared with the original one with rank(4719).  Hence, we conclude that the original model (with rank) is better for prediction.

\newpage
# Results

The model has decent prediction ability.  It should be noted that we only picked two positions to predict due to computation resource constraint.  In the pursue of creating this model, we have multiple route:

1) We have tried to include all positions.  However, the model took 5 hours to finish the MCMC simulation.  Therefore, we elected to include a smaller subset.
2) We have tried to include FanDuelSalary as a predictor.  However, that does not improve the model, and added sufficient time to compute - About 30 minutes to compute.  



# Contributions

- **Aaron Ray** (aaronwr2@illinois.edu)
- **Kiomars Nassiri** (nassiri2@illinois.edu)
- **Michael Chan** (mhchan3@illinois.edu)

# Reference

The analysis is inspired by the study presented in the article, **[Bayesian Hierarchical Modeling Applied to Fantasy Football Projections for Increased Insight and Confidence](http://srome.github.io/Bayesian-Hierarchical-Modeling-Applied-to-Fantasy-Football-Projections-for-Increased-Insight-and-Confidence/)**, by Scott Rome.

# Appendices

- Add data cleaning code here
